### 2020-08-11 Manuscript.Rmd
updated ms

# Evolution of gene expression across species and specialized zooids in Siphonophora {-}
Catriona Munro^1,2^,  Stefan Siebert^3^, Felipe Zapata^4^, Mark Howison^5^, Steven H.D. Haddock^6^, Casey W. Dunn^7^
\newline
\newline
^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI 02912, USA \newline
^2^ Current address: Collège de France, PSL Research University, CNRS, Inserm, Center for Interdisciplinary Research in Biology, 75005 Paris, France \newline
^3^ XXXXXX Insert Stefan address XXXXXXXXXXX \newline
^4^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA 90095, USA \newline
^5^ Watson Institute for International and Public Affairs, Brown University, Providence, RI 02912, USA \newline
^6^ Monterey Bay Aquarium Research Institute, Moss Landing, CA 95039, USA \newline
^7^ Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT 06520, USA

@@

We first considered the evolutionary change in expression along branches in the gene tree that correspond to branches in the species tree. These are identified as gene tree branches that have parent and child nodes that are both speciation events and that correspond to speciation events and branches in the species tree (Fig. \@ref(fig:methods-comp), step 5). This method enables the selection of specific branches within gene trees that are equivalent to branches within the species tree, and are thus comparable with one another across all gene trees. Unlike a strict 1:1 ortholog approach, this approach considers equivalent branches that are descended from speciation events, but that have more complex evolutionary histories. For example, due to deeper gene duplication events, gene trees often contain multiple branches that correspond to the same branch in the species tree (Fig. \@ref(fig:methods-comp), step 5). Our method allows us to consider all of these branches. Strict ortholog methods would discard some or all data impacted in this way by duplication events. Each speciation node in a gene tree is assigned an identifier that corresponds to the equivalent node in the species phylogeny. Each branch in the species tree is given a unique letter, as shown in figure \@ref(fig:changes), and the corresponding branches in gene trees are given the same letter. A simulated data set of random expression values was also obtained by using a Brownian motion model with empirically derived mean and standard-deviation values for each gene tree. Under this model, expression variance accumulates among linages on the gene tree as a function of time, and is used as a model of evolution under drift, as well as some forms of natural selection [@felsenstein1973maximum;@revell2008testing;@revell2008phylogenetic].

@@

The distribution of expression changes along branches in gene trees that correspond to equivalent branches in the species tree are shown in figure \@ref(fig:changes). The variance structure of the empirical data matches that of the simulated data (Fig. \@ref(fig:variance-structure)), suggesting that the size of standard deviation of change around the mean that is observed in figure \@ref(fig:changes) is a function of the underlying structure of the tree and the ancestral trait reconstruction methods that were used. High variance is especially pronounced in branches in the gene tree that correspond to K and L that have a shared parent node and lead to sister taxa *Nanomia bijuga* and *Agalma elegans* (Fig. \@ref(fig:variance-structure)). These two taxa are the only sister tip values that were sampled in this study. The mean values of change along branches do, however, differ from the Brownian model (Fig. \@ref(fig:changes)). Using linear models we find that expression change along branches in gene trees that correspond to branches in the species tree can be explained by significant differences among zooids/the pneumatophore, branches, and the interaction of the two (p < 2.2e-16, p = 2.319e-13, p < 2.2e-16, respectively, two-tailed anova). The Brownian model, based on simulated values on the gene trees under BM (p=0.4438, p=0.9829, p=0.4694), and also based on random reshuffling of changes on branches (p=0.31218, p=0.06787, p=0.0.54698), show that expression change along branches in gene trees that correspond to branches in the species tree cannot be explained by significant differences among zooids/the pneumatophore, branches, and the interaction of the two.

@@

Next, we were interested in branch level patterns of expression, focusing on whether change in expression across the same branch in a gene tree is correlated between different zooids. We investigated the correlation of changes across particular branches (that correspond to species tree branches) within gene trees among different zooids/the pneumatophore (Figs. \@ref(fig:correlations)). For change across the same branch within a gene tree that corresponds to a particular branch in the species tree, there is a positive correlation between pairwise comparisons of zooids/the pneumatophore. In particular, changes across all branches are highly correlated between developing and mature gastrozooids. Changes in expression across the same branch were also highly correlated between gastrozooids and palpons, especially so across branches corresponding to branch K (\@ref(fig:changes), top panel) in the species tree (branch leading to *Agalma elegans*).  Meanwhile, changes across all branches are much less correlated between any zooid/the pneumatophore and both the male or female gonodendra. These results suggest that with the exception of male and female gonodendra, in general changes in expression across gene tree branches are highly correlated among zooids - a large positive change in expression in one zooid across a gene tree branch will also typically see a large positive change in expression in another zooid across the same branch. By contrast, in the simulated Brownian motion dataset, we found that there was no correlation of changes across branches within gene trees among any zooid/tissue.

@@

Using this gene-tree centric approach, we were able to gain insights into general patterns of change across branches in gene trees that correspond to species tree branches. We found that most evolutionary changes across a branch in a gene tree are highly correlated among different zooids/tissues, with the exception of male and female gonodendra. We were also able to investigate the magnitude of these expression changes, and identify specific lineages and zooids with large increases (and decreases) in expression across gene tree branches. On average, gastrozooids (developing and mature), nectophores and pneumatophores show few large lineage specific changes in expression across gene tree branches. Mature palpons on average have large changes across branches, pointing to lineage specific changes in expression that are specific to this zooid. The large changes in palpon expression that we find along particular branches may be associated with lineage-specific adaptations. Unlike many other zooid types such as gastrozooids or gonophores that play a more conserved role within the colony, palpons have been observed to perform a wide range of functions in different siphonophore species. While many other zooid types such as gastrozooids or gonophores are thought to play relatively well-defined and conserved role within the colony, various functions have been ascribed to palpons in different siphonophore species. These include digestion, absorption, and egestion in *Nanomia bijuga* and *Forskalia* [@mackie1963fishing], and also defense in *Physophora hydrostatica* [@totton1965synopsis]. This latter case is particularly striking -- the palpons of *Physophora hydrostatica* vigorously attack objects placed in their vicinity [@totton1965synopsis]. The lack of knowledge about morphological traits specific to palpons and the sense that they don't have a consistent narrow function can give the sense that they are not as interesting as other zooids. The expression results presented here suggest that palpons should be viewed in a very different light -- they may have greater functional diversity than previously thought and are more fundamental to siphonophore biology than has been appreciated.

@@

We then took the mean TPM10K value for each gene across replicates of the same zooid/tissue within a species and applied a log transformation. Using gene trees with expression values for each gene within a species at the tips, maximum likelihood ancestral trait values were generated at the nodes using the anc.recon() function in `Rphylopars` assuming a Brownian model of evolution (Fig. \@ref(fig:methods-comp), step 2 & 3) [@goolsby2017rphylopars]. As not all zooids are present in all of the species, the trees were pruned down to the subset of tips with expression values for ancestral trait reconstructions. Node values were then added back to the unpruned tree with all of the reconstructed expression values. Change in expression was measured across a branch by taking the difference between a parent node and a child node, and then this difference is scaled by branch length (Fig. \@ref(fig:methods-comp), step 4).

@@

This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award). CM was also supported in part by a RI-EPSCoR Fellowship, NSF EPS-1004057. Analyses were conducted with computational resources and services at the Center for Computation and Visualization at Brown University, supported in part by the NSF EPSCoR EPS-1004057 and the State of Rhode Island. Analyses were also conducted with computational resources and services XXXXXXX Yale CCV

>>>>

# Evolution of gene expression across species and specialized zooids in Siphonophora {-}
Catriona Munro^1,2^,  Stefan Siebert^3^, Felipe Zapata^4^, Mark Howison^5^, Casey W. Dunn^6^
\newline
\newline
^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI 02912, USA \newline
^2^ Current address: Collège de France, PSL Research University, CNRS, Inserm, Center for Interdisciplinary Research in Biology, 75005 Paris, France \newline
^3^ XXXXXX Insert Stefan address XXXXXXXXXXX \newline
^4^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA 90095, USA \newline
^5^ Watson Institute for International and Public Affairs, Brown University, Providence, RI 02912, USA \newline
^6^ Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT 06520, USA

@@

We first considered the evolutionary change in expression along branches in the gene tree that correspond to branches in the species tree. These are identified as gene tree branches that have parent and child nodes that are both speciation events and that correspond to speciation events and branches in the species tree (Fig. \@ref(fig:methods-comp), step 5). This method enables the selection of specific branches within gene trees that are equivalent to branches within the species tree, and are thus comparable with one another across all gene trees. Unlike a strict 1:1 ortholog approach, this approach considers equivalent branches that are descended from speciation events, but that have more complex evolutionary histories. For example, due to deeper gene duplication events, gene trees often contain multiple branches that correspond to the same branch in the species tree (Fig. \@ref(fig:methods-comp), step 5). Our method allows us to consider all of these branches. Strict ortholog methods would discard some or all data impacted in this way by duplication events. Each speciation node in a gene tree is assigned an identifier that corresponds to the equivalent node in the species phylogeny. Each branch in the species tree is given a unique letter, as shown in figure \@ref(fig:changes), and the corresponding branches in gene trees are given the same letter. A simulated data set of random expression values was also obtained by using a Brownian Motion (BM) model with empirically derived mean and standard-deviation values for each gene tree. Under this model, expression variance accumulates among linages on the gene tree as a function of time, and is used as a model of evolution under drift, as well as some forms of natural selection [@felsenstein1973maximum;@revell2008testing;@revell2008phylogenetic].

@@

The distribution of expression changes along branches in gene trees that correspond to equivalent branches in the species tree are shown in figure \@ref(fig:changes). The variance structure of the empirical data matches that of the simulated data (Fig. \@ref(fig:variance-structure)), suggesting that the size of standard deviation of change around the mean that is observed in figure \@ref(fig:changes) is a function of the underlying structure of the tree and the ancestral trait reconstruction methods that were used. High variance is especially pronounced in branches in the gene tree that correspond to K and L that have a shared parent node and lead to sister taxa *Nanomia bijuga* and *Agalma elegans* (Fig. \@ref(fig:variance-structure)). These two taxa are the only sister tip values that were sampled in this study. The mean values of change along branches do, however, differ from the BM model (Fig. \@ref(fig:changes)). Using linear models we find that expression change along branches in gene trees that correspond to branches in the species tree can be explained by significant differences among zooids/the pneumatophore, branches, and the interaction of the two (p < 2.2e-16, p = 2.319e-13, p < 2.2e-16, respectively, two-tailed anova). The BM model, based on simulated values on the gene trees under BM (p=0.4438, p=0.9829, p=0.4694), and also based on random reshuffling of changes on branches (p=0.31218, p=0.06787, p=0.0.54698), show that expression change along branches in gene trees that correspond to branches in the species tree cannot be explained by significant differences among zooids/the pneumatophore, branches, and the interaction of the two.

@@

Next, we were interested in branch level patterns of expression, focusing on whether change in expression across the same branch in a gene tree is correlated between different zooids. We investigated the correlation of changes across particular branches (that correspond to species tree branches) within gene trees among different zooids/the pneumatophore (Figs. \@ref(fig:correlations)). For change across the same branch within a gene tree that corresponds to a particular branch in the species tree, there is a positive correlation between pairwise comparisons of zooids/the pneumatophore. In particular, changes across all branches are highly correlated between developing and mature gastrozooids. Changes in expression across the same branch were also highly correlated between gastrozooids and palpons, especially so across branches corresponding to branch K (\@ref(fig:changes), top panel) in the species tree (branch leading to *Agalma elegans*).  Meanwhile, changes across all branches are much less correlated between any zooid/the pneumatophore and both the male or female gonodendra. These results suggest that with the exception of male and female gonodendra, in general changes in expression across gene tree branches are highly correlated among zooids - a large positive change in expression in one zooid across a gene tree branch will also typically see a large positive change in expression in another zooid across the same branch. By contrast, in the simulated BM dataset, we found that there was no correlation of changes across branches within gene trees among any zooid/tissue.

@@

Using this gene-tree centric approach, we were able to gain insights into general patterns of change across branches in gene trees that correspond to species tree branches. We found that most evolutionary changes across a branch in a gene tree are highly correlated among different zooids/tissues, with the exception of male and female gonodendra. We were also able to investigate the magnitude of these expression changes, and identify specific lineages and zooids with large increases (and decreases) in expression across gene tree branches. On average, gastrozooids (developing and mature), nectophores and pneumatophores show few large lineage specific changes in expression across gene tree branches. Mature palpons on average have large changes across branches, pointing to lineage specific changes in expression that are specific to this zooid. The large changes in palpon expression that we find along particular branches may be associated with lineage-specific adaptations. Unlike many other zooid types such as gastrozooids or gonophores that play a more conserved role within the colony, palpons have been observed to perform a wide range of functions in different siphonophore species. These include digestion, absorption, and egestion in *Nanomia bijuga* and *Forskalia* [@mackie1963fishing], and also defense in *Physophora hydrostatica* [@totton1965synopsis]. This latter case is particularly striking -- the palpons of *Physophora hydrostatica* vigorously attack objects placed in their vicinity [@totton1965synopsis]. The lack of knowledge about morphological traits specific to palpons and the sense that they don't have a consistent narrow function can give the sense that they are not as interesting as other zooids. The expression results presented here suggest that palpons should be viewed in a very different light -- they may have greater functional diversity than previously thought and are more fundamental to siphonophore biology than has been appreciated.

@@

We then took the mean TPM10K value for each gene across replicates of the same zooid/tissue within a species and applied a log transformation. Using gene trees with expression values for each gene within a species at the tips, maximum likelihood ancestral trait values were generated at the nodes using the anc.recon() function in `Rphylopars` assuming a Brownian Motion model of evolution (Fig. \@ref(fig:methods-comp), step 2 & 3) [@goolsby2017rphylopars]. As not all zooids are present in all of the species, the trees were pruned down to the subset of tips with expression values for ancestral trait reconstructions. Node values were then added back to the unpruned tree with all of the reconstructed expression values. Change in expression was measured across a branch by taking the difference between a parent node and a child node, and then this difference is scaled by branch length (Fig. \@ref(fig:methods-comp), step 4).

@@

This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award). CM was also supported in part by a RI-EPSCoR Fellowship, NSF EPS-1004057. Analyses were conducted with computational resources and services at the Center for Computation and Visualization at Brown University, supported in part by the NSF EPSCoR EPS-1004057 and the State of Rhode Island. Analyses were also conducted with computational resources and services XXXXXXX Yale CCV. We would like to thank Steven Haddock, Lynne Christianson, and the whole crew of the Western Flyer for their help and expertise -- without them we would have been able to collect specimens. Additional thanks are due to Samuel Church for his assistance during one research cruise, and for productive conversations about comparative gene expression methods.
